Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-06-24 slides/slides.tex
slides update

Consider a \alert{co-algebraic} signature 
\[
\Sigma = \{id_n : n \to n,\sigma_{n,m} : n + m \to m + n,\mu : 1 \to 2, i : 1 \to 0\}
\]
and a family of terms $t$ such that every term is constructed from the generators above by combining them sequentially ($;$) and in parallel ($\otimes$).

@@

\begin{minted}{c++}
    enum CommutativeComonoids {
        "mu" = Mu,
        "i" = I,
        ";" = Compose([Id;2]),
        "parallel" = Parallel([Id;2]),
        "id" = Identity,
        "I" = Unit,
        "sym" = Sym    
    }
\end{minted}

@@

\begin{frame}
    However, the laws of monoidal category lead to an explosion of the number of e-classes
\end{frame}

\begin{frame}
    \begin{example}
        \vspace{1em}
        E-graph for
        \[
            (\mu\;;(\sigma_{1,1}\;;i \otimes id_{1})) \otimes (\mu\;;(id_{1}\otimes id_{1}))
        \]

        Before saturation
        \vspace{-2em}
        \begin{figure}
            \includegraphics[scale=0.4]{figures/egraph_before_saturation.pdf}
        \end{figure}
    \end{example}
\end{frame}

\begin{frame}
    \begin{example}
        \vspace{1em}
        E-graph for
        \[
            (\mu\;;(\sigma_{1,1}\;;i \otimes id_{1})) \otimes (\mu\;;(id_{1}\otimes id_{1}))
        \]
        After saturation
        \begin{figure}
            \includegraphics[width=\linewidth]{figures/dot_5.pdf}
        \end{figure}

@@

\begin{frame}
    \begin{example}
        \begin{minipage}{0.45\linewidth}
        Assuming $\mu \coloneqq $\raisebox{-1.5em}{\scalebox{0.6}{\begin{tikzpicture}
            \begin{pgfonlayer}{nodelayer}
                \node [style=round box] (0) at (0.5, 0.5) {$\mu$};
                \node [style=none] (1) at (0.5, 1.5) {};
                \node [style=none] (2) at (0, -0.5) {};
                \node [style=none] (3) at (1, -0.5) {};
                \node [style=none] (4) at (0.25, 0.25) {};
                \node [style=none] (5) at (0.75, 0.25) {};
            \end{pgfonlayer}
            \begin{pgfonlayer}{edgelayer}
                \draw (0) to (1.center);
                \draw [bend right=15] (4.center) to (2.center);
                \draw [bend left=15] (5.center) to (3.center);
            \end{pgfonlayer}
        \end{tikzpicture}}}
        and $i \coloneqq \;\;$\raisebox{-0.2em}{\scalebox{0.6}{\begin{tikzpicture}
            \begin{pgfonlayer}{nodelayer}
                \node [style=none] (0) at (3.5, 1) {};
                \node [style=vertex] (1) at (3.5, 0.5) {};
            \end{pgfonlayer}
            \begin{pgfonlayer}{edgelayer}
                \draw (0.center) to (1);
            \end{pgfonlayer}
        \end{tikzpicture}        
        }} a string diagrammatic version of
        \[ 
            (\mu\;;(\sigma_{1,1}\;;i \otimes id_{1})) \otimes (\mu\;;(id_{1}\otimes id_{1}))
        \] looks like
    \end{minipage}
        \hfill
    \begin{minipage}{0.45\linewidth}
        \begin{tikzpicture}
                \begin{pgfonlayer}{nodelayer}
                    \node [style=round box] (0) at (0.5, 0.5) {$\mu$};
                    \node [style=none] (1) at (0.5, 1.5) {};
                    \node [style=none] (2) at (0, -0.25) {};
                    \node [style=none] (3) at (1, -0.25) {};
                    \node [style=none] (4) at (0.25, 0.25) {};
                    \node [style=none] (5) at (0.75, 0.25) {};
                    \node [style=vertex] (6) at (1, -2.25) {};
                    \node [style=none] (7) at (1, -1.75) {};
                    \node [style=round box] (8) at (3.5, 0.5) {$\mu$};
                    \node [style=none] (9) at (3.5, 1.5) {};
                    \node [style=none] (10) at (3, -0.25) {};
                    \node [style=none] (11) at (4, -0.25) {};
                    \node [style=none] (12) at (3.25, 0.25) {};
                    \node [style=none] (13) at (3.75, 0.25) {};
                    \node [style=none] (16) at (0.5, -1) {};
                    \node [style=none] (17) at (0, -1.75) {};
                    \node [style=none] (18) at (1, -1.75) {};
                    \node [style=none] (19) at (0, -2.25) {};
                    \node [style=none] (20) at (3, -2.25) {};
                    \node [style=none] (21) at (4, -2.25) {};
                \end{pgfonlayer}
                \begin{pgfonlayer}{edgelayer}
                    \draw (0) to (1.center);
                    \draw [bend right=15] (4.center) to (2.center);
                    \draw [bend left=15] (5.center) to (3.center);
                    \draw (6) to (7.center);
                    \draw (8) to (9.center);
                    \draw [bend right=15] (12.center) to (10.center);
                    \draw [bend left=15] (13.center) to (11.center);
                    \draw [in=30, out=-90] (3.center) to (16.center);
                    \draw [in=90, out=-135] (16.center) to (17.center);
                    \draw [in=315, out=90] (18.center) to (16.center);
                    \draw [in=-90, out=150] (16.center) to (2.center);
                    \draw (17.center) to (19.center);
                    \draw (10.center) to (20.center);
                    \draw (11.center) to (21.center);
                \end{pgfonlayer}
            \end{tikzpicture}
    \end{minipage}
    \vfill
    Which is already saturated with respect to the built-in equations
    \end{example} 
\end{frame}

>>>>

Consider a \alert{co-algebraic} signature 
\[
\Sigma = \{id_0 : 0 \to 0, id_{1} : 1 \to 1, \sigma_{1,1} : 1 + 1 \to 1 + 1,\mu : 1 \to 2, i : 1 \to 0\}
\]
and a family of terms $t$ such that every term is constructed from the generators above by combining them sequentially ($;$) and in parallel ($\otimes$).

@@

\begin{minted}{c++}
    enum CommutativeComonoids {
        "mu" = Mu([Id;1]),
        "i" = I([Id;1]),
        ";" = Compose([Id;2]),
        "parallel" = Parallel([Id;2]),
        "id" = Identity([Id;1]),
        "I" = Unit,
        "sym" = Sym([Id;2])   
    }
\end{minted}

@@

\begin{frame}
    However, because of the laws of monoidal category this does not scale very well
\end{frame}

\begin{frame}
    \small
    \begin{example}
        \vspace{1em}
        E-graph for
        \[
            (\mu\;;(\sigma_{1,1}\;;i \otimes id_{1})) \otimes (\mu\;;(id_{1}\otimes id_{1}))
        \]

        Before saturation
        \vspace{-3em}
        \begin{figure}
            \includegraphics[scale=0.4]{figures/egraph_before_saturation.pdf}
        \end{figure}
    \end{example}
\end{frame}

\begin{frame}
    \small
    \begin{example}
        \vspace{1em}
        E-graph for
        \[
            (\mu\;;(\sigma_{1,1}\;;i \otimes id_{1})) \otimes (\mu\;;(id_{1}\otimes id_{1}))
        \]
        After saturation
        \begin{figure}
            \includegraphics[width=0.9\linewidth]{figures/dot_5.jpeg}
        \end{figure}

@@

\begin{frame}
    \begin{example}
        \begin{minipage}{0.45\linewidth}
        Assuming $\mu \coloneqq $\raisebox{-1.5em}{\scalebox{0.6}{\begin{tikzpicture}
            \begin{pgfonlayer}{nodelayer}
                \node [style=round box] (0) at (0.5, 0.5) {$\mu$};
                \node [style=none] (1) at (0.5, 1.5) {};
                \node [style=none] (2) at (0, -0.5) {};
                \node [style=none] (3) at (1, -0.5) {};
                \node [style=none] (4) at (0.25, 0.25) {};
                \node [style=none] (5) at (0.75, 0.25) {};
            \end{pgfonlayer}
            \begin{pgfonlayer}{edgelayer}
                \draw (0) to (1.center);
                \draw [bend right=15] (4.center) to (2.center);
                \draw [bend left=15] (5.center) to (3.center);
            \end{pgfonlayer}
        \end{tikzpicture}}}
        and $i \coloneqq \;\;$\raisebox{-0.2em}{\scalebox{0.6}{\begin{tikzpicture}
            \begin{pgfonlayer}{nodelayer}
                \node [style=none] (0) at (3.5, 1) {};
                \node [style=vertex] (1) at (3.5, 0.5) {};
            \end{pgfonlayer}
            \begin{pgfonlayer}{edgelayer}
                \draw (0.center) to (1);
            \end{pgfonlayer}
        \end{tikzpicture}        
        }} a string diagrammatic version of
        \[ 
            (\mu\;;(\sigma_{1,1}\;;i \otimes id_{1})) \otimes (\mu\;;(id_{1}\otimes id_{1}))
        \] looks like
    \end{minipage}
        \hfill
    \begin{minipage}{0.45\linewidth}
        \begin{tikzpicture}
                \begin{pgfonlayer}{nodelayer}
                    \node [style=round box] (0) at (0.5, 0.5) {$\mu$};
                    \node [style=none] (1) at (0.5, 1.5) {};
                    \node [style=none] (2) at (0, -0.25) {};
                    \node [style=none] (3) at (1, -0.25) {};
                    \node [style=none] (4) at (0.25, 0.25) {};
                    \node [style=none] (5) at (0.75, 0.25) {};
                    \node [style=vertex] (6) at (1, -2.25) {};
                    \node [style=none] (7) at (1, -1.75) {};
                    \node [style=round box] (8) at (3.5, 0.5) {$\mu$};
                    \node [style=none] (9) at (3.5, 1.5) {};
                    \node [style=none] (10) at (3, -0.25) {};
                    \node [style=none] (11) at (4, -0.25) {};
                    \node [style=none] (12) at (3.25, 0.25) {};
                    \node [style=none] (13) at (3.75, 0.25) {};
                    \node [style=none] (16) at (0.5, -1) {};
                    \node [style=none] (17) at (0, -1.75) {};
                    \node [style=none] (18) at (1, -1.75) {};
                    \node [style=none] (19) at (0, -2.25) {};
                    \node [style=none] (20) at (3, -2.25) {};
                    \node [style=none] (21) at (4, -2.25) {};
                \end{pgfonlayer}
                \begin{pgfonlayer}{edgelayer}
                    \draw (0) to (1.center);
                    \draw [bend right=15] (4.center) to (2.center);
                    \draw [bend left=15] (5.center) to (3.center);
                    \draw (6) to (7.center);
                    \draw (8) to (9.center);
                    \draw [bend right=15] (12.center) to (10.center);
                    \draw [bend left=15] (13.center) to (11.center);
                    \draw [in=30, out=-90] (3.center) to (16.center);
                    \draw [in=90, out=-135] (16.center) to (17.center);
                    \draw [in=315, out=90] (18.center) to (16.center);
                    \draw [in=-90, out=150] (16.center) to (2.center);
                    \draw (17.center) to (19.center);
                    \draw (10.center) to (20.center);
                    \draw (11.center) to (21.center);
                \end{pgfonlayer}
            \end{tikzpicture}
    \end{minipage}
    \vfill
    Which is already `saturated' with respect to the built-in equations
    \end{example} 
\end{frame}

A hypergraph form of the above example is

\[
\scalebox{0.6}{
    \tikzfig{figures/hypergraph_with_interfaces}
}    
\]
